---
title: "Framework for PBPK modeling in GNU MCSim"
author: "Nan-Hung Hsieh"
date: "`r Sys.Date()`"
output: 
  html_document:
  fig_caption: yes  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 0 Overview

In this tutorial we want to know,

<strong>

- How to write Physiologically based pharmacokinetic (PBPK) model?

- How to write input file and conduct simulation?

- How to analyze *GNU MCSim* output with *R*?

</strong>

Here, we'll use tetrachloroethylene (PERC) and ethylbenzene (EB) as examples to illustrate how to perform straight simulation in *GNU MCSim* and analyze the output result in *R*.

---

```{r, include=FALSE}
# source("MCSim/function.R")
# library(tidyverse)
```

## 1 Create PBPK model

This section illustrates how to build a PBPK model in *GNU MCSim*. In the beginning, use comments (`#`) to annotate your code. Here is the example that you can use to comment on your model file.

- Brief description of this model
- What is the version number
- When you create this model
- Summarize the unit used in this model

Then, follow the next seven steps to build the PBPK model.

### 1.1 Define the state variables [`States{}`](https://www.gnu.org/software/mcsim/mcsim.html#Global-variable-declarations)

States are variables for which a first-order differential equation is defined in the `Dynamics{}` section. Remember, when naming the variables, be sure to let all variable names begin with a capital letter followed by meaningful lower case subscripts. In `States{}` section, each variable needs to separate by comma `,`. In addition, remember to provide the unit that is used for the state variables. 

### 1.2 Define the outputs variables [`Outputs{}`](https://www.gnu.org/software/mcsim/mcsim.html#Global-variable-declarations)

Outputs are dependent model variables (obtainable at any time as analytical functions of the states, inputs or parameters) that do not have dynamics. They must receive assignments in either the `Dynamics` or `CalcOutputs` sections. When constructing the model, we need to have an output variable for checking mass balance. Then, assign the variable that we are interested in or have an experiment to make the comparison, such as the blood concentration.

### 1.3 Define the input variable(s) [`Inputs{}`](https://www.gnu.org/software/mcsim/mcsim.html#Global-variable-declarations)

Inputs are variables independent of the other variables, but eventually varying with time (for example an exposure concentration to a chemical). *GNU MCSim* has various functions to define the different exposure types as:

- PerDose(): # specifies a periodic input of constant

    `PerDose(<magnitude>, <period>, <initial-time>, <exposure-time>);`

- PerExp(): # specifies a periodic exponential input.

    `PerExp(<magnitude>, <period>, <initial-time>, <decay-constant>);`

- PerTransit(): #models a delayed input mechanism (e.g., gut absorption) 

    `PerTransit(<magnitude>, <period>, <initial-time-in-period>, <decay-constant>, <number-of-input-compartments>);`  
    
- NDoses(): # specifies a number of stepwise inputs of variable magnitude and their starting times
    
    `NDoses(<n>, <list-of-magnitudes>, <list-of-initial-times>);`

- Spikes(): # specifies a number of instantaneous inputs of variable magnitude and their exact times of occurrence.
    
    `Spikes(<n>, <list-of-magnitudes>, <list-of-times>);`

### 1.4 Parameters of the model

Here we need to define the "default" value of input parameters. Unlike previous variables that need to use curly brackets to group these variables, the input parameters can be put in any places without the restriction. In addition, don't forget to put a semicolon (`;`) after the definition of each parameter. If the input parameter doesn't have a default value, just put a semicolon after the name of the parameter or used an arbitrary number. 

The PBPK parameters always include the parameters that can describe absorption, distribution (e.g., partition coefficient), metabolism (e.g., Michaelis-Menten constant), and elimination (e.g., rate Constant). These parameters can be classified to chemical-specific (e.g., molecular weight) and anatomy (e.g., body weight, blood flow, tissue weight/volume) parameter. In addition, the unit conversion factor can define in this part. 

The parameters of the PBPK model might include: 

- Unit conversion factor 

- Exposure modeling parameters

- Chemical-specific parameters

- Physiological parameters

- Pharmacokinetic parameters

- Scale parameters (no default value)

### 1.5 Define the parameter initialization and scaling [`Initialize{}`](https://www.gnu.org/software/mcsim/mcsim.html#Model-initialization)

This section is used to scale specific model parameters and resolve dependencies between parameters. Generally, the scaling involves a change of units or conversion from percentage to actual units.

### 1.6 Define the dynamics of the simulation [`Dynamics{}`](https://www.gnu.org/software/mcsim/mcsim.html#Dynamics-section)

The equations given in this section will be called by the integration routines at each integration step. It includes specification of differential equations.

The derivative of a state variable is defined using the `dt()` operator, as shown here:

`dt(state-variable) '=' constant-value-or-expression ';'`

### 1.7 Define the output calculations [`CalcOutputs{}`](https://www.gnu.org/software/mcsim/mcsim.html#Output-calculations) 

In this section, the equations are given in this section will be called by the simulation program at each output time specified by a `Print()` or `PrintStep()` statement. In this way, output computations are done efficiently, only when values are to be saved. Here is the example to assign the variables in `CalcOutputs{}` section. Again, don't forget the semicolon and put the keyword `End.` at the end of model. 

---

## 2 Create input/simulation file

Same as model file, we highly recommend to use comments (`#`) to annotate your code in the beginning. This comment might include:

- Brief description of this input file
- What model file that is used to conduct the simulation
- Others...

### 2.1 Setting integrator [`Integrate()`](https://www.gnu.org/software/mcsim/mcsim.html#Integrate_0028_0029-specification)

*GNU MCSim* provides three types of integrator that include `Lsodes`, `Cvodes`, and `Euler`. Here, we used `Lsodes` (Livermore Solver for Ordinary Differential Equation) as a primary solver, which was originated from the SLAC Fortran library. The `Lsodes` function uses adaptive numerical methods to advance a solution to a system of ordinary differential equations one time-step, given values for the variables Y and X. It solves the initial value problem for stiff or non-stiff systems of ordinary differential equations (ODE).

The syntax for `Lsodes` is: `Integrate(Lsodes, <rtol>, <atol>, <method>);`

where `<rtol>` is a scalar specifying the relative error tolerance for each integration step. The scalar `<atol>` specifies the absolute error tolerance parameter. Those tolerances are used for all state variables. The estimated local error for a state variable y is controlled so as to be roughly less (in magnitude) than rtol*|y| + atol. Thus the local error test passes if, for each state variable, either the absolute error is less than <atol>, or the relative error is less than `<rtol>`. Set `<rtol>` to zero for pure absolute error control, and set `<atol>` to zero for pure relative error control. Caution: actual (global) errors may exceed these local tolerances, so choose them conservatively. Decreasing the tolerance leads to more accurate results but at the cost of significant increase in time taken.

The `<method>` flag should be 0 (zero) for non-stiff differential systems and 1 or 2 for stiff systems. You should try flag 0 or 1 and select the fastest for equal accuracy of output unless insight from your system leads you to choose one of them a priori. If you specify <method> 2 you should provide the Jacobian of your differential system. The good starting point for <atol> and <rtol> is about 1e-6. The default integration method in `Lsodes` with parameters 1e-5, 1e-7 and 1.

### 2.2 Setting paramerer(s)

For the simple simulation, we can assign the given parameter values in the input file to replace the default values in the model file. The parameter can be set in the `simulation{}` section as a local setting or outside the section as a global setting. Same as the parameter setting in the model file, use `=` to define the given value of parameters and put `;` after the definition.

### 2.3 Setting simulation [`simulation{}`]()

After the global specifications, we can define the local parameters. In addition, we need to define the output through [Print()](https://www.gnu.org/software/mcsim/mcsim.html#Print_0028_0029-specification) or [PrintStep()](https://www.gnu.org/software/mcsim/mcsim.html#PrintStep_0028_0029-specification) functions. The arguments of `Print()` and `PrintStep()` are the comma-separated list of variable names (at least one and up to MAX_PRINT_VARS, which is set to 10 by default) and followed by given time-points (limited by the available memory at run time) as:

`Print(<identifier1>, <identifier2>, ..., <time1>, <time2>, ...);`

`PrintStep(<identifier1>, <identifier2>, ..., <start-time>, <end-time>, <time-step>);`

**Note:** The output variables cannot receive assignments in simulation input files.

After finish the model- and input-file, we can use `mcsim()` to run the simulation. This is the example of use `mcsim()` to solve Tetrachloroethylene (PERC) PBPK model

```{r}
mcsim(model = "perc.model.R", input = "perc.in.R")
```

### 2.4 Working with the R package deSolve

In addition to use MCSim to solve ODE, there is an alternative way to solve ODE. The integrators provided by **deSolve** are improved implementations of the lsode family of integrators used by GNU MCSim. They provide a few more options than *GNU MCSim*.
such as the functions that solve initial value problems of a system of partial differential equations ('PDE'), of differential algebraic equations ('DAE'), and of delay differential equations. They provide an interface to the FORTRAN functions 'lsoda', 'lsodar', 'lsode', 'lsodes' of the 'ODEPACK' collection, to the FORTRAN functions 'dvode', 'zvode' and 'daspk' and a C-implementation of solvers of the 'Runge-Kutta' family with fixed or variable time steps. However, *GNU MCSim* is the fastest option in intensive simulation.

---

## 3 Analyze *GNU MCSim* output with *R*

*GNU MCSim* doesn't include a comprehensive tool to analyze and visualize the simulation result. The best way to do further data analysis and visualization is to use *R*. This instruction will provide some basic R functions that can use to learn and understand how to analyze *GNU MCSim* output with *R*. 

### 3.1 Loading output

The simulation output is written under tab-delimited text file with default name `sim.out`. It is located in the directory. You can check your file through files tab on the bottom-right panel in RStudio. Or use `file.exists()` to check the file as,

```{r}
# file.exists()
```

If you can find your output file in your working directory, then you can us R function `read.delim()` to read it. Since the first line of the output file is used to describe the number of simulation, it can't be used in the formal data structure. Therefore, we need to assign `skip = 1` to ignore this line. Then run the following code and check your result:

```{r}
# read.delim(file = , skip = )
```

Through this function, we can only read the content in the output file, but we can’t use the data unless we assign it to a variable. Therefore, we need to assign a variable for the output to help us analyze it. The assigned variables will be temporarily stored in memory. 

### 3.2 Assign variables

Use the assignment operator `<-` to create new variables.

```{r}
# out <- read.delim(file = , skip = )
```

Once a variable is created, we can use the variable name to refer to the simulation data it was assigned. To view the variable in RStudio, you can interact with it in the top-right panel. Or, use `View()` to check it:

```{r, eval=F}
# View()
```

### 3.3 Manipulating output

Now, our data is loaded into R. We can start doing further analysis with them. First, let’s check what type it is:

```{r}
# class()
```

The output tells us that it’s a "data frame". This type of data structure constructed by the specific number of column (variable) and row (observation). We can see the dimensions of the data frame with the function `dim()`: 

```{r}
# dim()
```

Or use `nrow()` and `ncol()` to check the information:

```{r}
# nrow()
# ncol()
```

Also, the `names()` can be used to get and set the names of variables

```{r}
# names()
```

Some basic functions can be used to summarize the result, such as `mean()`, `min()`, `max()`. Here is an example to find the maximum concentration in blood:  

```{r}
# max()
```

Here, the `$` is used to assign the specific variable that was stored in the object. Also, you can quickly check all maximum or minimum concentration through `apply`:

```{r}
# apply(out, MARGIN = , FUN = )
```

You can also use `which()` to find the value as:

```{r}
# which(out$C_ven == max(out$C_ven))
# which(out$C_ven == min(out$C_ven))
```

But, the `which()` only give us the location, not time point. Therefore, we need to use this location to find the corresponding time point by:

```{r}
# which_row <- which(out$C_ven == max(out$C_ven))
# out$Time[]
```

The following code can let you set the variable name:

```{r}
# names(out) <- c(, , )
```

Check again,

```{r}
# names(out)
```

If you only want to change the specific variable, you can use:

```{r}
# names(out)[] <- 
# names(out)
```

The unit conversion is the cruical part in data manipulation as well. Here we want to transfer the unit from mg/l to mol/l,

```{r}
# MW <- 165.822 # g/mol
# out$C_blood_mol <- out$C_blood / MW * 1000 # mg/l -> mol/l
```

Finally, if you prefer to use Excel, you can use `write.csv()` to transfer your object to Comma-separated values (csv) file, then you can use Excel to open it and do additional analysis.

```{r}
# write.csv(, file =, row.names = )
```

### 3.4 Plotting

Visualizing the output data is an important work to communicate the simulation result. After the data manipulation, we can use `plot()` to create the PK diagram as:

```{r}
# plot(x = , y = )
```

Usually, the basic assignment cannot provide high quality and comprehensive figure that can further use in our report or published paper. Therefore, we need to use additional assignments such as `xlab`, `ylab`, `main` to add the description or modification.

```{r}
# plot(x = , y= , 
#     type = , col = , xlab =  ylab = ,
#     log = , main = )
```

Here is an example to plot the two simulation results (concentration of exhaled air and blood) in the same plot. However, we can find that the range of y-axis doesn't cover all simulation points.

```{r}
# plot()
```

To solve this problem, we can customize the range x-axis with the `range()` to define the range of x-axis as: 

```{r}
# y_rng <- range()
# plot()
```

In addition, the relationship of concentration in blood and air can further be visualized through this way,

```{r}
# plot()
```

Finally, you can use `pdf()` to export the plotting result to portable document format (pdf) as:

```{r}
# pdf("perc_pk.pdf")
# dev.off()
```

## 4 One-compartment model

### 4.1 Modeling and simulation 

This is an example of one-compartment model. Here we need to use model and input files of `one.model.R` and `one.in.R` to coduct the simulation. First, let's check the output 

```{r}
# out.1cpt <- mcsim(model = , input = )
# out.1cpt
```

Since the default function cannot load the output file with multi simulation result under ideal format to perform further analysis, we can use our built-in function `readsims()` to re-analyze the output object.

```{r}
# exp_1 <- readsims(, exp = )
# exp_2 <- readsims(, exp = )
```

Plot the result of `oral_input` 

```{r}
# plot(, type = , col = , xlab = , ylab = )
# lines(, type = )
```

and `C_central`.

```{r}
#
#
```

### 4.2 Exercise

In `perc.in.R`, add the second simulation group that has an exposure concentration of 144 ppm. Besides, add `Pct_metabolized` in both simulation sections.

```{r}
# out.perc <- mcsim(model = "", input = "")
```

Load the data to two objects,

```{r}
# perc_1 <- readsims(, exp = 1)
# perc_2 <- readsims(, exp = 2)
```

Plot time vs exhaled air

```{r}
# plot(, , type = , log = , col = , 
#     xlab = , ylab = )
# lines()
# legend("topright", legend = , col = , lty = )
```

Plot time vs blood concentration

```{r}
#
#
#
```

Plot time vs Percent metabolized

```{r}
#
#
#
```

## 5 Example: Ethylbenzene (EB) PBPK model

In the EB-PBPK example, we generated three exposure scenarios in the input file, which were (1) 8-hr continuous exposure to 100 ppm EB, (2) 8-hr continuous exposure to 1 mg/m<sup>3</sup> EB, and (3) 4-hr continuous exposure to 100 ppm EB. For the first and second scenarios, we'll compare the EB concentration in arterial blood (`Cart`),  pulmonary (`Cvipu`),    and venous blood (`Cvtot`). The third case will be used to compare with the experimental data from the published study (Tardif et al., 1996).

### 5.1 Modeling and simulation 

Same as PERC-PBPK case, we firstly use `mcsim()` to simulate the pharmacokinetics under the default parameter setting for the three given scenarios.

```{r}
# out.EB <- mcsim(model = , input = )
```

Use `readsims()` to re-analyze the output object:

```{r}
#EB_1 <- readsims(, exp = ) 
#EB_2 <- readsims(, exp = ) 
#EB_3 <- readsims(, exp = ) 
```

According to the above information, we can find that we have a relative "long" output format than the previous case for PERC. Therefore, we used `head()` and `tail()` to check the output and prevent the long report. The assignment `n` can use any integer to change the length of data.

```{r}
#head()
#tail(, n = )
```

Same as the previous case, use `apply()` to find the maximum concentration in scenario 1 and 2.

```{r}
# apply(, MARGIN = , FUN = )
```

For case 3, we need to further transfer the original unit (mol/L) to **mg/L** that can compare with the result in the published paper.

```{r}
# MW <- 106.16 # g/mol
# EB_3$Cvtot_mg <-  # mol/L -> mg/L 
# apply(, MARGIN = , FUN = )
```

After the data manipulation, we can further visualize our simulation result to compare the EB concentration sin arterial blood (`Cart`),  pulmonary (`Cvipu`),    and venous blood (`Cvtot`) under 8-hr continuous exposure for

(1) 100 ppm, and  

```{r}
#plot(EB_1$Time, EB_1$Cvipu, col = , type = , main = , 
#     xlab = , ylab = )
# lines(, , col = )
# lines(, , col = )
# legend("topright", legend = , col = , lty = )
```

(2) 1 mg/m<sup>3</sup>

```{r}
#
#
#
#
```

Finally, we want to compare the simulation result with experiment data as:

```{r}
# data_x <- c(4.02, 4.5, 5, 5.5, 6)
# data_y <- c(1.93, 1.29, 0.87, 0.55, 0.38)
# sd <- c(0.15, 0.15, 0.24, 0.11, 0.05)
# sd.up <- data_y + sd
# sd.dn <- data_y - sd
# plot(data_x, data_y, main = "Exposure: 100 ppm", xlab = "Time (hr)", ylab = "Concentration (mg/L)",
#     xlim = c(0, 6), ylim = c(0, 2.5))
# lines(x = , y =)
# arrows(x0 = , y0 = , x1 = , y1 = , code = , length = , angle = )
# legend("topright", legend = , lty = , pch = )
```

According to our simulation and experiment data, we found that the current parameter setting can't provide the suitable prediction with the experiment data. Therefore, we might need to calibrate our parameter setting to perfom further modeling and simulation.

### 5.2 Exercise

#### 5.2.1 Run EB-PBPK model under 100 ppm exposure for 4 hours and plot the time-course of blood concentration from 0 to 6 hour.

1. Modeling
```{r}
# out <- mcsim(model = , input = )
```

2. Data manipulate

3. Visualize
```{r}
# plot()
```

#### 5.2.2 Estimate the steady-state of arterial and venous blood concentrations associated with EB exposures (0.1 ppm to 1000 ppm).

1. Modeling
```{r}
# out <- mcsim(model = , input = )
```

2. Data manipulate
```{r}

```

3. Visualize
```{r}
# plot()
# lines()
# legend()
```

#### 5.2.3 Construct the relationships for the estimated inhalation exposure level and the fraction of EB metabolized after 8-hr continuous exposure. In addition, estimate the percentage metabolized from liver, lung, and richly perfused tissue.

1. Modeling

```{r}
# out <- mcsim(model = , input = )
```

2. Data manipulate
```{r}

```

3. Visualize
```{r}
# plot()
# lines()
# lines()
# lines()
```

#### 5.2.4 Add additional exposure routes include oral ingestion in the EB-PBPK model and estimate the Cmax and Tmax after received a single gavage dose of 180 mg/kg.

1. Modeling
```{r}
# out <- mcsim(model = , input = )
```

2. Data manipulate
```{r}
# C_max
# T_max
```

3. Visualize
```{r}
# plot()
```

### Reference

Bois, F.Y., Gelman A., Jiang J., Maszle D.R., Zeise L., Alexeef G. (1996). [Population toxicokinetics of tetrachloroethylene](https://link.springer.com/article/10.1007/s002040050284).   
Chiu, WA and Bois, Y. (2006). [Revisiting the population toxicokinetics of tetrachloroethylene](https://link.springer.com/article/10.1007/s00204-006-0061-9).  
Gelman A., Bois, F.Y., Jiang J. (1996). [Physiological Pharmacokinetic Analysis Using Population Modeling and Informative Prior Distributions](https://www.tandfonline.com/doi/abs/10.1080/01621459.1996.10476708?journalCode=uasa20).  
[GNU MCSim user manual](https://www.gnu.org/software/mcsim/mcsim.html)  
Soetaert, K., et al. (2018). [deSolve: Solvers for Initial Value Problems of Differential Equations ('ODE', 'DAE', 'DDE')](https://cran.r-project.org/web/packages/deSolve/index.html).    
Tardif, R., Charest-Tardif, G., and Brodeur, J. (1996). [Comparison of the influence of binary mixtures versus a ternary mixture of inhaled aromatic hydrocarbons on their blood kinetics in the rat](https://link.springer.com/article/10.1007/s002040050292).  

